Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-05-06 Master Thesis/bibliography.tex
Add Kato to bibliography
Add items to bibliography
Add introduction

\begin{thebibliography}{99999}

\bibitem[Sp96]{Sp96}
	M. Spiess (1996)
	\textit{An arithmetic proof of Pop's Theorem concerning Galois groups of function fields over number fields}.
    J. reine angew. Math. \textbf{478}, 107\---126.

>>>>

\begin{thebibliography}{999999}

\bibitem[EP05]{EP05}
    A.J. Engler, A. Prestel
    \textit{Valued Fields}.
    Springer Monographs in Mathematics (2005).

\bibitem[Ka86]{Ka86}
    K. Kato
    \textit{A Hasse principle for two dimensional global fields, with an appendix by J.-L. Colliot Thélène}.
    J. reine angew. Math. \textbf{366} (1986), 142\---183.

\bibitem[Ne69]{Ne69}
    J. Neukirch
    \textit{Kennzeichnung der $p$-adischen und der endlichen algebraischen Zahlkörper}.
    Invent. Math. \textbf{6} (1969), 296\---314.

\bibitem[Ne77]{Ne77}
    J. Neukirch
    \textit{Über die absoluten Galoisgruppen algebraischer Zahlkörper}. Asterisque, \textbf{41\---42}, Soc. Math. France, Paris (1977), 67\---79.

\bibitem[Po90]{Po90}
    F. Pop
    \textit{On the Galois theory of function fields of one variable over number fields}.
    J. reine angew. Math. \textbf{406} (1990), 200\---218.

\bibitem[Po94]{Po94}
    F. Pop
    \textit{On Grothendieck's conjecture of birational anabelian geometry}.
    Ann. Math. \textbf{138} (1994), 145\---182.

\bibitem[Po95]{Po95}
    F. Pop
    \textit{On Grothendieck's conjecture of birational anabelian geometry II}, Preprint (1995).

\bibitem[Sp96]{Sp96}
	M. Spiess
	\textit{An arithmetic proof of Pop's Theorem concerning Galois groups of function fields over number fields}.
    J. reine angew. Math. \textbf{478} (1996), 107\---126.

\bibitem[Uc76]{Uc76}
    K. Uchida
    \textit{Isomorphisms of Galois groups}.
    J. Math. Soc. Japan \textbf{28} (1976), no. 4, 617\---620.

\bibitem[Uc77]{Uc77}
    K. Uchida
    \textit{Isomorphisms of Galois groups of algebraic function fields}.
    Ann. Math. (2) \textbf{106} (1977), no. 3, 589\---598.

\bibitem[ZS60]{ZS60}
    O. Zariski, P. Samuel
    \textit{Commutative Algebra, Volume II} (1960). 
    Van Nostrand.

### 2017-05-06 Master Thesis/introduction.tex
Update notations
Add n-th roots of unity to list of notations
Add introduction

\begin{theorem}[\textit{Pop}]\label{thm:main-result}
Let $K,K'$ be two finite field extensions of $\mathbb{Q}(T)$. If there is an isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups, then there exists a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$ with $\sigma(g)=\varphi^{-1}g\varphi$ for all $g\in G_K$. In particular $\varphi$ induces a field isomorphism $K'\to K$.
\end{theorem}

\begin{theorem}[\textit{Neukirch-Uchida}]\label{thm:neukirch-uchida}
Let $k,k'$ be two global fields. Then any isomorphism $G_k\to G_{k'}$ is induced by a unique isomorphism $k'^{\text{s}}\to k^\text{s}$.  In particular, if $k,k'$ have isomorphic absolute Galois groups, they are isomorphic as fields.
\end{theorem}

@@

$\begin{array}{ll}
\mathbb{N} & \text{natural numbers} \\
\mathbb{Z} & \text{integers}\\
\mathbb{Q} & \text{rational numbers}\\
\mathbb{R} & \text{real numbers}\\
\mathbb{C} & \text{complex numbers}\\
\\
\overline{K} & \text{algebraic closure of $K$}\\
K^\text{s} & \text{separable closure of $K$}\\
K^\text{t} & \text{absolute inertia field of $K$}\\
K_v & \text{henselization of $K$ with respect to $v$}\\
G_K & \text{absolute Galois group of $K$}\\
\operatorname{Gal}(L/K) & \text{Galois group of $L/K$}\\
\operatorname{char}(K) & \text{characteristic of $K$}\\
\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}
\end{array}$

>>>>

Given an algebraic variety $X$, we can ask ourselves how much information about $X$ is contained in the knowledge of its étale fundamental group $\pi_1^{\text{ét}}(X)$. Especially interesting are schemes which should be reconstructible solely from its étale fundamental group. If $X$ is the spectrum of a field $K$, the étale fundamental group of $X$ is just its absolute Galois group $G_K$.

Finite fields have an abelian absolute Galois group. In fact, they all have the same absolute Galois group $\hat{\mathbb{Z}}$, therefore not much information can be retrieved from these groups. The study of these types of questions belong in the area of \textit{anabelian geometry}. This term refers to the fact that the less abelian $\pi_1^\text{ét}(X)$ is, the more information it carries about $X$.

Number fields stand in stark contrast to finite fields. Two number fields $k$, $k'$ with isomorphic Galois groups $G_k$, $G_{k'}$ are isomorphic as fields as well. It was J. Neukirch who investigated this question for the first time in the end of the 1960s (see \cite{Ne69}, \cite{Uc76}, \cite{Ne77}, \cite{Uc77}). Together with K. Uchida, their studies resulted in the following theorem for global fields:

\begin{theorem}[\textit{Neukirch-Uchida}]\label{thm:neukirch-uchida}
Let $k,k'$ be two global fields. Then any isomorphism $\sigma: G_k\to G_{k'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi: k'^{\text{s}}\to k^\text{s}$, i.e: 
\[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_k\] 
In particular $\varphi$ induces a field isomorphism $k'\to k$.
\end{theorem}

This means that global fields should be reconstructible from their Galois groups. Furthermore, the Galois group $G_k$ also contains important arithmetic information from $k$.

It is believed that spectra of finitely generated fields are the building blocks of the anabelian world, which led Grothendieck to the conjecture that Theorem \ref{thm:neukirch-uchida} should also hold for arbitrary finitely generated fields $k$ and $k'$ of characteristic $0$.

This has been proved by F. Pop (see \cite{Po90}, \cite{Po94}, \cite{Po95}) using model theory. In this thesis, we want to follow M. Spiess \cite{Sp96} to prove the statement for finitely generated fields with transcendence degree $1$ over $\mathbb{Q}$, using a purely arithmetical approach and a direct generalization of the original proof of Theorem \ref{thm:neukirch-uchida}:

\pagebreak

\begin{theorem}[\textit{Pop}]\label{thm:main-result}
Let $K,K'$ be two finite field extensions of $\mathbb{Q}(T)$. Then any isomorphism $\sigma:G_K\to G_{K'}$ between their absolute Galois groups is induced by a unique field isomorphism $\varphi:\overline{K'}\to\overline{K}$, i.e: 
\[\sigma(g) =\varphi^{-1}g\varphi \quad \text{for all }g\in G_k\]
In particular $\varphi$ induces a field isomorphism $K'\to K$.
\end{theorem}

@@

$\begin{array}{ll}
\mathbb{N} & \text{natural numbers} \\
\mathbb{Z} & \text{integers}\\
\mathbb{Q} & \text{rational numbers}\\
\mathbb{R} & \text{real numbers}\\
\mathbb{C} & \text{complex numbers}\\
\mathbb{F}_q & \text{finite field with $q$ elements}\\
\mathbb{Z}_p & \text{$p$-adic integers}\\
\hat{\mathbb{Z}} & \text{profinite completion of $\mathbb{Z}$}\\
\\
\overline{K} & \text{algebraic closure of $K$}\\
K^\text{s} & \text{separable closure of $K$}\\
K^\text{t} & \text{absolute inertia field of $K$}\\
K_v & \text{henselization of $K$ with respect to $v$}\\
G_K & \text{absolute Galois group of $K$}\\
\operatorname{Gal}(L/K) & \text{Galois group of $L/K$}\\
\operatorname{char}(K) & \text{characteristic of $K$}\\
\operatorname{trdeg}(L/K) & \text{transcendence degree of $L/K$}\\
\mu_n & \text{$n$-th roots of unity in a separable closure}
\end{array}$

### 2017-05-06 Master Thesis/thesis.tex
Remarks are numbered
Add working title
Update title
Add proof for 4.5
Solve TeX warnings
Add introduction

\documentclass[11pt, a4paper, openany]{memoir}
\usepackage{fontspec}
\usepackage[english]{babel}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{paralist, tikz}
\usetikzlibrary{cd, babel}
\usepackage[inner=4cm, outer=3cm, top=3cm, bottom=3cm]{geometry}
\usepackage{hyperref}
\hypersetup{bookmarksopen=true, pdfstartview=FitH}
\linespread{1.1}
\author{Yichuan Shen}
\title{Title of the Master Thesis}
\makeindex

@@

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem*{definition}{Definition}
\newtheorem*{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem{step}{Step}

@@

\makeatletter
\begin{center}
\vspace*{0cm}
\begin{large}
Universität Heidelberg\\
Fakultät für Mathematik und Informatik\\
Studiengang Mathematik\\
\vspace{8mm}
\end{large}
\vfill
\begin{large}
\textbf{MASTER THESIS}\\
\end{large}
\vspace{10mm}
\begin{huge}\@title \end{huge}\\
\vspace{10mm}
by \textsc{yichuan shen}\\
\vspace{3cm}
\vfill
Supervisor: \textsc{prof. dr. alexander schmidt}\\
\today
\end{center}
\makeatother

@@

\vspace*{0cm}
\vfill
\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature and other professional sources. This thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
\vspace{7mm}\\
\rule{7cm}{0.4pt}
\clearpage

@@

\input{synopsis.tex}
\input{bibliography.tex}

\printindex
\end{document}

>>>>

\documentclass[11pt, a4paper, openany]{memoir}
\usepackage{fontspec}
\usepackage[english]{babel}
\usepackage{amsmath, amsfonts, amssymb, amsthm}
\usepackage{paralist, tikz}
\usetikzlibrary{cd, babel}
\usepackage[inner=4cm, outer=3cm, top=3cm, bottom=3cm]{geometry}
\usepackage{hyperref}
\hypersetup{bookmarksopen=true, pdfstartview=FitH}
\linespread{1.1}
\author{Yichuan Shen}
\title{Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields}

@@

\theoremstyle{plain}
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem*{definition}{Definition}
\newtheorem*{example}{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{step}{Step}

@@

\makeatletter
\begin{center}
\vspace*{0cm}
\begin{large}
Universität Heidelberg\\
Fakultät für Mathematik und Informatik\\
Studiengang Mathematik\\
\vspace{8mm}
\end{large}
\vfill
\begin{large}
\textbf{MASTER THESIS}\\
\end{large}
\vspace{10mm}
\linespread{2}\selectfont
\begin{huge}\@title \end{huge}\\
\linespread{1.1}\selectfont
\vspace{10mm}
by \textsc{yichuan shen}\\
\vspace{3cm}
\vfill
Supervisor: \textsc{prof. dr. alexander schmidt}\\
\today
\end{center}
\makeatother

@@

\vspace*{0cm}
\vfill
\paragraph{Declaration.} I declare that I carried out this master thesis independently, and only with the cited sources, literature, and other professional sources. This thesis was not used in the same or in a similar version to achieve an academic grading or is being published elsewhere.\\
\vspace{7mm}\\
\rule{7cm}{0.4pt}
\clearpage

@@

\input{synopsis.tex}
\input{bibliography.tex}
\end{document}
